---
title: "DSL Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DSL Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This document describes every field in the ksTable JSON DSL.

## Top-level structure

```json
{
  "schema_version": "1.0",
  "table_spec": {
    "id":         "<string>",
    "title":      "<string>",
    "parameter":  { ... },
    "statistics": { ... },
    "groups":     { ... },
    "layout":     { ... }
  }
}
```

| Field | Required | Description |
|-------|----------|-------------|
| `schema_version` | No | DSL version string (currently `"1.0"`) |
| `table_spec.id` | No | Unique identifier for the table |
| `table_spec.title` | No | Human-readable title |
| `table_spec.parameter` | **Yes** | Map of parameters (see below) |
| `table_spec.statistics` | **Yes** | Map of statistics (see below) |
| `table_spec.groups` | **Yes** | Grouping/stratification |
| `table_spec.layout` | **Yes** | Row/column structure |

---

## `parameter`

A named map where each key is a parameter identifier and the value describes
a variable in the input data.

```json
"parameter": {
  "<param_id>": {
    "variable": "<column_name>",
    "label":    "<display_label>",
    "nested":   { "<child_id>": { "variable": "<col>", "label": "<lbl>" } }
  }
}
```

| Field | Required | Description |
|-------|----------|-------------|
| `variable` | **Yes** | Column name in the input `data` frame. Must be a valid R identifier. |
| `label` | No | Display label for this parameter row. Defaults to `variable`. |
| `nested` | No | One child parameter for `hierarchical` row structure (SOC → PT). |

### Example — simple

```json
"parameter": {
  "age": { "variable": "AGE", "label": "Age (years)" }
}
```

### Example — hierarchical

```json
"parameter": {
  "soc": {
    "variable": "AESOC",
    "label":    "",
    "nested": {
      "pt": { "variable": "AEDECOD" }
    }
  }
}
```

---

## `statistics`

A named map where each key is a statistic identifier and the value describes
how to compute and display it.

```json
"statistics": {
  "<stat_id>": {
    "fun":    "<function_name>",
    "args":   { "<arg>": <value>, ... },
    "format": { "type": "<type>", ... }
  }
}
```

| Field | Required | Description |
|-------|----------|-------------|
| `fun` | **Yes** | Name of the calc function in the evaluation environment. Must be a valid R identifier. |
| `args` | No | Extra named arguments passed to `fun` (see below). |
| `format` | No | Format spec that converts the raw value to a display string. Defaults to `as.character()`. |

### `args`

Extra arguments appended to the `fun(variable, ...)` call in the generated script.
Keys must be valid R identifiers; values are converted to R literals.

```json
"args": {
  "na.rm":  true,
  "probs":  0.25,
  "digits": 2,
  "method": "type7"
}
```

Generated: `quantile(AGE, na.rm = TRUE, probs = 0.25, digits = 2, method = "type7")`

| JSON value | R literal |
|------------|-----------|
| `true` | `TRUE` |
| `false` | `FALSE` |
| `null` | `NULL` |
| `1.5` | `1.5` |
| `"text"` | `"text"` |

### `format`

Determines how `.value_raw` (the raw output of `fun`) is converted to the
display string `.value`. Every statistic has a formatter; the default when
`format` is omitted is `as.character()`.

#### No format spec (default)

```json
"n": { "fun": "count" }
```

Generated: `.value = as.character(.value_raw)`

#### `"type": "sprintf"`

```json
"age_mean": {
  "fun":    "mean",
  "args":   { "na.rm": true },
  "format": { "type": "sprintf", "pattern": "%.1f" }
}
```

Generated: `.value = sprintf("%.1f", .value_raw)`

#### `"type": "custom"`

Used when the calc function returns a **named list**. The format function
receives the list element and returns a character string.

```json
"mean_sd": {
  "fun":    "mean_sd",
  "format": { "type": "custom", "fun": "format_mean_sd" }
}
```

Generated: `.value = vapply(.value_raw, format_mean_sd, character(1L))`

The `format.fun` value is looked up in the evaluation environment — it is a
plain function name, not a list entry.

#### `"type": "template"`

Interpolates a [glue](https://glue.tidyverse.org/) template against the named
list returned by the calc function. Requires `glue` to be installed.

```json
"mean_sd": {
  "fun":    "mean_sd",
  "format": { "type": "template", "pattern": "{mean} ({sd})" }
}
```

Generated: `.value = vapply(.value_raw, function(.x) glue::glue_data(.x, "{mean} ({sd})"), character(1L))`

#### `"type": "ksformat"`

Applies a [ksformat](https://github.com/crow16384/ksformat) VALUE format to
the raw value.

```json
"trt": {
  "fun":    "identity",
  "format": { "type": "ksformat", "format_name": "trt_fmt" }
}
```

Generated: `.value = ksformat::fput(.value_raw, "trt_fmt")`

---

## `groups`

Defines how the input data is stratified into output columns.

```json
"groups": {
  "by":                     ["TRT01P", "SEX"],
  "include_missing_levels": true,
  "format": {
    "TRT01P": "trt_format"
  }
}
```

| Field | Required | Description |
|-------|----------|-------------|
| `by` | **Yes** | Array of column names to group by. Each must be a valid R identifier. |
| `include_missing_levels` | No | If `true`, emit `group_by(..., .drop = FALSE)` so factor levels with zero observations still appear as rows. Requires the grouping columns to be factors with the desired levels set (use `kst_apply_metadata()`). Default `false`. |
| `format` | No | Optional documentation map from column name to ksformat format name. **Not applied by `kst_compile()`.** Pass the same mapping as `format_map` to `kst_extract_metadata()` / `kst_apply_metadata()` before eval. |

### include_missing_levels

To use `include_missing_levels: true`, ensure grouping columns are factors
with all desired levels before evaluating compiled code:

```{r missing-levels}
library(ksTable)

# Extract factor levels (pass groups.format mapping as format_map when using ksformat)
meta <- kst_extract_metadata(adsl, "TRT01P",
                             format_map = list(TRT01P = "trt_format"))

# Re-level the data
adsl_levelled <- kst_apply_metadata(adsl, meta)

code <- kst_compile(spec)
env <- new.env(parent = environment())
env$data <- adsl_levelled
result <- eval(parse(text = code), envir = env)
```

---

## `layout`

```json
"layout": {
  "row_structure":    "parameter_stat",
  "column_structure": "groups"
}
```

| Field | Required | Description |
|-------|----------|-------------|
| `row_structure` | **Yes** | How rows are structured (see below). |
| `column_structure` | No | Currently only `"groups"` is supported. |

### `row_structure` values

| Value | Description | Output rows |
|-------|-------------|-------------|
| `"parameter_stat"` | One row per parameter × statistic combination. | `.param`, `.stat` columns. |
| `"hierarchical"` | Parent rows + child rows (e.g., SOC → PT). Requires exactly one `nested` parameter. | `.parent`, `.is_child`, `.row_label`, `.stat` columns. |

For `hierarchical`, rows are sorted by `.parent` ascending, then parents
before children (`.is_child = FALSE < TRUE`), then by `.row_label`
alphabetically.

---

## Output tibble structure

### `parameter_stat`

```
.param       character  parameter label
.stat        character  statistic key
<group1>     character  formatted value for group 1
<group2>     character  formatted value for group 2
...
```

### `hierarchical`

```
.parent      character  parent row label (e.g. SOC name)
.is_child    logical    FALSE = parent row, TRUE = child row
.row_label   character  display label (same as .parent for parent rows)
.stat        character  statistic key
<group1>     character  formatted value for group 1
...
```

The `.parent`, `.is_child`, `.row_label`, `.stat` columns are consumed
by ksTFL to apply indentation and group headers during rendering.

---

## Complete example — efficacy table

Statistics using only built-in R functions, all with `na.rm = TRUE`:

```json
{
  "schema_version": "1.0",
  "table_spec": {
    "id": "efficacy",
    "title": "Change from Baseline in Systolic BP",
    "parameter": {
      "chg": { "variable": "CHG", "label": "Change from Baseline (mmHg)" }
    },
    "statistics": {
      "n":      { "fun": "sum",      "args": { "na.rm": true } },
      "mean":   { "fun": "mean",     "args": { "na.rm": true },
                  "format": { "type": "sprintf", "pattern": "%.1f" } },
      "sd":     { "fun": "sd",       "args": { "na.rm": true },
                  "format": { "type": "sprintf", "pattern": "%.2f" } },
      "median": { "fun": "median",   "args": { "na.rm": true },
                  "format": { "type": "sprintf", "pattern": "%.1f" } },
      "min":    { "fun": "min",      "args": { "na.rm": true },
                  "format": { "type": "sprintf", "pattern": "%.1f" } },
      "max":    { "fun": "max",      "args": { "na.rm": true },
                  "format": { "type": "sprintf", "pattern": "%.1f" } }
    },
    "groups": {
      "by": ["TRT01P", "AVISIT"],
      "include_missing_levels": false
    },
    "layout": {
      "row_structure":    "parameter_stat",
      "column_structure": "groups"
    }
  }
}
```

No user-defined calc or format functions are needed — all six statistics use
standard R functions with `na.rm = TRUE` from the JSON spec, and `sprintf`
for display formatting.

---

## Security (SR-1)

All JSON fields that appear in generated code are validated before emission:

* **Identifier fields** (`variable`, `fun`, `format.fun`, `by` items, `args` keys):
  validated against `^[A-Za-z.][A-Za-z0-9._]*$`. Any value containing `)`,
  `;`, spaces, or other metacharacters is rejected with an informative error.

* **String literal fields** (`label`, `pattern`): `\` and `"` are escaped
  before embedding in generated R string literals.

`kst_validate_spec()` performs these checks and returns all errors before any
code is generated.
