---
title: "Getting Started with ksTable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ksTable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

**ksTable** reads a declarative JSON table specification and generates a plain,
human-readable dplyr/tidyr R script. Running that script against your data
produces a formatted tibble ready for rendering with
[ksTFL](https://github.com/crow16384/ksTFL).

```
JSON spec  →  kst_compile()  →  R script  →  eval()  →  tibble  →  ksTFL
```

Key design principles:

* **Declarative**: describe *what* you want, not how to compute it.
* **Plain-R output**: the generated script is ordinary dplyr — paste it into
  the console, step through it, save it to a file.
* **No infrastructure**: calc and format functions are resolved from your
  environment just like any other R function call.
* **Isolated execution**: intermediate objects (`.chunks`, `.long`) stay in a
  private environment and never pollute your workspace.

---

## Installation

```{r install, eval = FALSE}
remotes::install_github("crow16384/ksTable")
```

```{r library}
library(ksTable)
```

---

## Function contracts

ksTable uses two distinct function types, both defined by the user in their
own environment.

### Calc functions

Referenced by the `"fun"` field in a statistic.  Receive a **column vector**
(one group's data), return a raw **numeric** or **named list**.
*No string formatting here.*

```{r calc-fns}
count   <- function(data) sum(!is.na(data))          # → integer

mean_sd <- function(data) list(                       # → named list
  mean = mean(data, na.rm = TRUE),
  sd   = sd(data,   na.rm = TRUE)
)

count_n <- function(data) length(data)               # → integer
```

### Format functions

Referenced by `"format": { "type": "custom", "fun": "..." }`.  Receive the
raw value from the corresponding calc function, return a single
**character** string.

```{r format-fns}
format_mean_sd <- function(x) sprintf("%.1f (%.2f)", x$mean, x$sd)
```

Built-in R functions (`median`, `sum`, `mean`, …) can be used directly as
calc functions.  Extra arguments — such as `na.rm = TRUE` — are specified in
the JSON spec via the optional `"args"` field (see below).

---

## Example 1: Demographics table

### Synthetic data

```{r data}
set.seed(42)
n <- 480L

adsl <- data.frame(
  USUBJID = sprintf("%03d-%04d", rep(1:3, each = n %/% 3L), seq_len(n)),
  AGE     = round(rnorm(n, 45, 12)),
  SEX     = sample(c("M", "F"), n, replace = TRUE),
  TRT01P  = rep(c("Placebo", "Drug A", "Drug B"), each = n %/% 3L),
  stringsAsFactors = FALSE
)
```

### JSON spec

Three statistics are requested:

* **n** — a user-defined `count()` function.
* **mean_sd** — a user-defined `mean_sd()` returning a named list, formatted
  by a user-defined `format_mean_sd()`.
* **median** — the built-in `median()` with `na.rm = TRUE` supplied via `"args"`.

```{r demog-json}
demog_spec <- '{
  "schema_version": "1.0",
  "table_spec": {
    "id": "demographics_age",
    "title": "Age by Treatment Group",
    "parameter": {
      "age": { "variable": "AGE", "label": "Age (years)" }
    },
    "statistics": {
      "n": {
        "fun":   "count"
      },
      "mean_sd": {
        "fun":    "mean_sd",
        "format": { "type": "custom", "fun": "format_mean_sd" }
      },
      "median": {
        "fun":   "median",
        "args":  { "na.rm": true }
      }
    },
    "groups": {
      "by": ["TRT01P"],
      "include_missing_levels": false
    },
    "layout": {
      "row_structure":    "parameter_stat",
      "column_structure": "groups"
    }
  }
}'
```

### Inspect the generated script

`kst_compile()` returns a character string you can read before running it:

```{r compile}
code <- kst_compile(demog_spec)
cat(code)
```

```
.chunks <- list()

.chunks[[1L]] <- data |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarize(
    .value_raw  = count(AGE),
    .groups     = "drop"
  ) |>
  dplyr::mutate(
    .value      = as.character(.value_raw),
    .param      = "Age (years)",
    .stat = "n",
    .value_raw  = NULL
  )

.chunks[[2L]] <- data |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarize(
    .value_raw  = list(mean_sd(AGE)),
    .groups     = "drop"
  ) |>
  dplyr::mutate(
    .value      = vapply(.value_raw, format_mean_sd, character(1L)),
    .param      = "Age (years)",
    .stat = "mean_sd",
    .value_raw  = NULL
  )

.chunks[[3L]] <- data |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarize(
    .value_raw  = median(AGE, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  dplyr::mutate(
    .value      = as.character(.value_raw),
    .param      = "Age (years)",
    .stat = "median",
    .value_raw  = NULL
  )

.long <- dplyr::bind_rows(.chunks)

tidyr::pivot_wider(
  .long,
  id_cols     = c(.param, .stat),
  names_from  = c(TRT01P),
  values_from = .value
)
```

### Generate the table

```{r generate-demog}
code <- kst_compile(demog_spec)
env <- new.env(parent = environment())
env$data <- adsl
result <- eval(parse(text = code), envir = env)
print(result)
```

```
# A tibble: 3 × 5
  .param      .stat   `Drug A`     `Drug B`     Placebo
1 Age (years) n       160          160          160
2 Age (years) mean_sd 44.9 (11.37) 44.9 (11.84) 44.5 (11.96)
3 Age (years) median  45           44.5         44
```

---

## Example 2: Adverse events table (hierarchical)

The `hierarchical` row structure produces parent rows (SOC) with child rows
(PT) nested beneath them.

### Additional data

```{r ae-data}
set.seed(7)
n_ae   <- 300L
ae_ids <- sample(adsl$USUBJID, n_ae, replace = TRUE)
ae_soc <- sample(c("Cardiac disorders", "Gastrointestinal disorders"),
                 n_ae, replace = TRUE, prob = c(0.4, 0.6))
ae_pt  <- ifelse(
  ae_soc == "Cardiac disorders",
  sample(c("Myocardial infarction", "Angina pectoris"), n_ae, replace = TRUE),
  sample(c("Nausea", "Diarrhea"), n_ae, replace = TRUE)
)
ae <- data.frame(
  USUBJID = ae_ids,
  TRT01P  = adsl$TRT01P[match(ae_ids, adsl$USUBJID)],
  AESOC   = ae_soc,
  AEDECOD = ae_pt,
  stringsAsFactors = FALSE
)
```

### JSON spec

```{r ae-json}
ae_spec <- '{
  "schema_version": "1.0",
  "table_spec": {
    "id": "adverse_events",
    "title": "Adverse Events by SOC and PT",
    "parameter": {
      "soc": {
        "variable": "AESOC",
        "label":    "",
        "nested": {
          "pt": { "variable": "AEDECOD" }
        }
      }
    },
    "statistics": {
      "n": {
        "fun":   "length",
        "args":  {}
      }
    },
    "groups": {
      "by": ["TRT01P"],
      "include_missing_levels": false
    },
    "layout": {
      "row_structure":    "hierarchical",
      "column_structure": "groups"
    }
  }
}'
```

### Generate the table

```{r generate-ae}
ae_code <- kst_compile(ae_spec)
ae_env <- new.env(parent = environment())
ae_env$data <- ae
ae_result <- eval(parse(text = ae_code), envir = ae_env)
print(ae_result)
```

```
# A tibble: 6 × 7
  .parent             .is_child .row_label           .stat Drug A Drug B Placebo
1 Cardiac disorders   FALSE     Cardiac disorders    n           36     39     34
2 Cardiac disorders   TRUE      Angina pectoris      n           14     21     21
3 Cardiac disorders   TRUE      Myocardial infarction n          22     18     13
4 Gastrointestinal d… FALSE     Gastrointestinal d…  n           73     64     54
5 Gastrointestinal d… TRUE      Diarrhea             n           39     31     35
6 Gastrointestinal d… TRUE      Nausea               n           34     33     19
```

The `.parent` and `.is_child` columns identify the hierarchy; ksTFL uses them
to apply indentation and header formatting during rendering.

---

## Step-through debugging

Run the compiled script in `new.env(parent = environment())`.
For interactive debugging, set up the environment manually and inspect
intermediate objects:

```{r debug}
code <- kst_compile(demog_spec)

env       <- new.env(parent = environment())   # calc/format functions visible via parent
env$data  <- adsl

# Run the whole script at once
result <- eval(parse(text = code), envir = env)

# Or paste individual lines into the console to step through:
#   eval(parse(text = ".chunks <- list()"), envir = env)
#   eval(parse(text = ".chunks[[1L]] <- data |> ..."), envir = env)
#   env$.chunks[[1L]]   # inspect first chunk

# All intermediates are available for inspection
ls(env, all.names = TRUE)
# → ".chunks"  ".long"  "data"

env$.long   # the assembled long-format table before pivoting
```

---

## Using built-in R functions

Any function available in the calling environment can be used as a calc
function.  Pass extra arguments via `"args"` in the JSON spec:

```json
"statistics": {
  "n":      { "fun": "sum",      "args": { "na.rm": true  } },
  "median": { "fun": "median",   "args": { "na.rm": true  } },
  "q25":    { "fun": "quantile", "args": { "probs": 0.25, "na.rm": true } },
  "sd":     { "fun": "sd",       "args": { "na.rm": true  } }
}
```

Generated code (for `median`):

```r
.value_raw = median(AGE, na.rm = TRUE)
```

`"args"` values are converted to R literals:

| JSON type | R literal |
|-----------|-----------|
| `true` / `false` | `TRUE` / `FALSE` |
| `null` | `NULL` |
| number | numeric literal |
| string | double-quoted character |

---

## Validation

`kst_validate_spec()` checks the spec without generating any code. It is
useful for early feedback and for testing JSON produced by tooling or users.

```{r validate-ok}
result <- kst_validate_spec(demog_spec)
result$valid    # TRUE
result$errors   # character(0)
```

```{r validate-bad}
bad_spec <- '{"schema_version":"1.0","table_spec":{"parameter":{},
               "statistics":{},"groups":{"by":["TRT"]},"layout":{}}}'
result <- kst_validate_spec(bad_spec)
result$valid
# [1] FALSE
result$errors
# [1] "Missing required field: /table_spec/layout/row_structure"
```

Injection attempts are rejected at the identifier level (SR-1):

```{r validate-inject}
inject <- '{"schema_version":"1.0","table_spec":{"parameter":{
              "p":{"variable":"AGE); system(\\"rm -rf /\\""}},
              "statistics":{"n":{"fun":"count"}},"groups":{"by":["TRT"]},
              "layout":{"row_structure":"parameter_stat"}}}'
result <- kst_validate_spec(inject)
result$valid
# [1] FALSE
result$errors
# [1] "'AGE); system(\"rm -rf /\"' is not a valid R identifier ..."
```

---

## Performance

Compilation is ~130–150 µs per call (pure R string assembly). At that rate,
compiling is negligible overhead even in loops or Shiny apps.

```{r benchmark}
system.time(for (i in 1:1000L) kst_compile(demog_spec))
```

---

## Next steps

* See `vignette("dsl_reference")` for the complete JSON schema reference.
* The generated script can be saved to a `.R` file and used independently of
  the package — useful for audit trails in regulated environments.
* Set the parent of your eval environment (for example,
  `new.env(parent = environment())`) when you need precise control over
  function lookup scope.
